---
title: "afp"
author: "Robert Schnitman"
date: "June 23, 2018"
output: html_document
---
***Robert Schnitman***  
***2018-06-23***  
**Recommended Citation:**  
  **Schnitman, Robert (2018). afp v0.0.0.1. https://github.com/robertschnitman/afp **

```{r setup, include=FALSE}
pkg_names <- c('do.bind', 'mapreduce', 'telecast')
pkg_paths <- paste0('r/', pkg_names, '.r')

invisible(lapply(pkg_paths, source)) # load packages.
```

## Outline
0. Installation
1. Introduction
2. `do.bind()`
3. `mapreduce()`
4. `telecast()`
5. Conclusion

## 0. Installation
```{r s0, eval = FALSE}
## Dependencies: R >= 3.5

# install.packages("devtools")
devtools::install_github("robertschnitman/afp")

```

## 1. Introduction

The `afp` package--*Applied Functional Programming*--provides functionals to simplify iterative processes in R. The Base R `*apply()` family, `purrr` library, and [Julia programming language](https://julialang.org/) are the principal influences. The former two contain tools essential for functional programming that minimize the need to incorporate loops and increase code brevity; however, there is inelegance with respect to specific situations. 

For example, to map a function and consequently bind rows or columns, `purrr` splits the decision into two functions rather than within one: `map_dfr()` and `map_dfc()`, both of which only output to data frames--while this feature is as intended, they nonetheless omit the possibility of a matrix when such a data type is preferred. Additionally, to reduce the results of a mapping, one must encase `Map()`/`lapply()` in `Reduce()`, while [Julia blends the routine into `mapreduce()`](https://docs.julialang.org/en/v0.6.1/stdlib/collections/#Base.mapreduce-NTuple{4,Any}).

As such, `afp` exists to supplement the functionals in Base R, `purrr`, and others to support efficient and concise programming.

The following sections provide examples for the primary functions: `do.bind()`, `mapreduce()`, and `telecast()`.

## 2. `do.bind()`

When executing `lapply()` to manipulate subsets of data, calling `rbind()` or `cbind()` within `do.call()` is common to fuse the transformed partitions. While the implementation is possible on one line by way of the `do.call(*bind, lapply(x, f))` form, the readability and intended concision decreases as the complexity of the anonymous function increases. Even if the `lapply()` portion is stored in an object before hand, the intention of `do.call()` is not clear until `*bind` is stated. To minimize these issues, `do.bind()` wraps this process and clarifies its purpose: *bind the results of the given function*.

One may obtain similar results with `map_dfr()`/`map_dfc()` from `purrr`; but the output would always result in a data frame rather than the possibility of a matrix. Additionally, the binding rows or columns must be done in different functions, whereas it can be defined within `do.bind()`.

There are three required parameters in this function: `f`, `x`, and `m`--respectively the function, collection (e.g. data frame), and margin (rbind/cbind designation). If `m = 1` (the default), the results are combined row-wise; `2` for column-wise. A fourth parameter `...` passes to `do.call()`. 

To note, the naming of this function is to be consistent with the previously mentioned function.

### EXAMPLE - `do.bind()` \#1: Store subset-contingent lm() coefficients in a matrix.
```{r s2-1}
split1  <- split(mtcars, mtcars$gear)
adhoc1  <- function(s) {coef(lm(mpg ~ disp + wt + am, s))}
output1 <- do.bind(adhoc1, split1) # == do.call(rbind, lapply(split1, adhoc1)).
output1                            # Rownames indicate subset.

```

### EXAMPLE - `do.bind()` \#2: Compute the median ozone-temperature ratio for each given month.
```{r s2-2}
airquality2 <- na.omit(airquality) # NA values exist in airquality.
split2      <- split(airquality2, airquality2$Month)
adhoc2      <- function(x, y) {median(x/y)}
output2     <- do.bind(function(s) with(s, adhoc2(Ozone, Temp)), split2, 2) 
output2 # == do.call(cbind, lapply(split2, function(s) with(s, adhoc2(Ozone, Temp))))

```

## 3. `mapreduce()`

Base R has functionals that output a list by way of `lapply()` and `Map()` (among others), while `Reduce()` diminishes given elements in a consecutive manner until a single result remains. While these functions are relatively straightforward to combine (depending on the functions being passed), R does not inherently possess a singular function to accomplish this operation [unlike Julia with `mapreduce()`](https://docs.julialang.org/en/v0.6.1/stdlib/collections/#Base.mapreduce-NTuple{4,Any}).

The three required parameters are `f`, `o`, and `x` ("fox", collectively). A fourth parameter `y` can be used when `f` imposes additional arguments. The last parameter `...` passes to `Reduce()`.

In turn, `mapreduce()` in `afp` offers a simplified Julia-equivalent with a multivariate option to streamline the intended procedure in R.

### EXAMPLE - `mapreduce()` \#1: Element-wise divide manipulated matrices.
```{r s3-1}
matrixl <- list(A = matrix(c(1:9), 3, 3), B = matrix(10:18, 3, 3), C = matrix(19:27, 3, 3))
output1 <- mapreduce(function(x) x^2 + 1, `/`, matrixl) # == Reduce(`/`, Map(function(x) x^2 + 1, matrixl))
output1

```

### EXAMPLE - `mapreduce()` \#2: A case of multiple arguments.
```{r s3-2}
# Using the same list from the previous example...
output2 <- with(matrixl, mapreduce(function(i, j, k) i*j - k, `/`, A, list(B, C)))
output2

```

## 4. `telecast()`

`Map()`/`mapply()` from Base R executes functions pairwise when given multiple data objects, as do `map2()`/`pmap()` from `purrr`. While beneficial in its own right, said functions cannot concisely map over datasets *independently* of each other, which would be useful for storing disparate information into a single list.

Inspired by [`broadcast()` from Julia](https://docs.julialang.org/en/v0.6.1/manual/arrays/#Broadcasting-1), `telecast()` essentially wraps `mapply()` within `lapply()` to achieve this outcome.

The two functions (both required) are `f` and `l`, respectively a function and list. The third parameter `as.vector`, which is optional, converts the output to a vector if set to `TRUE` (and thus will resemble the output from `rapply()`); by default, it is `FALSE` for a list format.

### EXAMPLE - `telecast()` \#1: Apply means to each variable for all datasets in a stored list.
```{r s4-1}
l <- list(mc = mtcars, aq = airquality, lcs = LifeCycleSavings)

mean.nr <- function(x) mean(x, na.rm = TRUE) # airquality has NA values.
output1 <- telecast(mean.nr, l) 
output1 # Compare: lapply(l, function(x) mapply(mean.nr, x))

```

### EXAMPLE - `telecast()` \#2: Iteratively reduce each variable for all datasets in a list and store in a vector.
```{r s4-2}
# With the same list in the previous example...

red.div <- function(y) Reduce(`/`, na.omit(y))
output2 <- telecast(red.div, l, as.vector = TRUE) 
output2 # Compare: rapply(l, red.div)
```

## 5. Conclusion

The functions discussed and demonstrated will be improved on a continuous basis to (1) minimize repetitive iterative processing and (2) emphasize code efficiency and brevity. New functions to be added based on feasibility and future needs as appropriate.

*End of Document*